An Accurate Model for SARS-CoV-2 Pooled RT-PCR Test Errors

### 2021-09-14 paper.tex
Amit's remarks

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in the fight against the COVID-19 pandemic. RT-PCR
testing can be performed on individuals suspected of infection and
hence break potential transmission chains. Moreover, in some
countries, RT-PCR testing has become a routine procedure of
surveillance, testing incoming air passengers in airports
\cite{PoolingAirports, TestingAirportsJapan}, schools students
\cite{PoolSchool}, or even the entire adult population
\cite{MassPooling}.

As such, the need for large-scale testing has resulted in the
application of techniques to increase the throughput of RT-PCR tests
\cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman, MatrixPooling,
  LionDorfman, CherifReview}. These techniques, often termed
\emph{pooling}, or group testing, have originated in the 1940's in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroughly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

@@

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID, CherifReview}. Consequently,
studies assumed that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool,
  CherifReview}, as well as in more specific, SARS-CoV-2 focused
studies \cite{Simplistic1, Simplistic2}. In this study we challenge
this commonly made assumption and demonstrate how using a more accurate
probabilistic model affects the estimation of false-negative rates for
Dorfman pooling.

Several consequences arise from our analysis: Previously estimated
optimal Dorfman pool size \cite{OptimalDorfmanPool} should be
revisited in the case of SARS-CoV-2, as it is influenced by the
prevalence of infection. The error rates of other pooling schemes in
use should also be revisited \cite{BayesianDorfman, Kim}. Another
consequence of our results is that studies that seek precise error
models should be conducted for other pathogens as well, since the
standard error model could prove flawed for other
pathogens. Specifically, all assumptions should be scrutinized,
including (e.g.) the independence of tests, an extremely widespread
assumption in the group testing literature \cite{Kim,
  OptimalDorfmanPool} which dates back to Dorfman's original paper
\cite{DorfmanOriginal}. Finally, new methods should be developed in
order to take advantage of our (and other) more realistic models.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume a more
realistic model: amplification of viral RNA succeeds or fails for each
sample independently. Furthermore, according to \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool either. For lack of
data pointing otherwise, we incorporate this assumption into our model
with a small modification. We do assume that a false amplification can
occur only once per pool. However, we also assume false amplification
is independent of any other correct amplification. Specifically, it is
possible that every correct amplification fails \emph{and} an
erroneous one occurs simultaneously. This assumption is somewhat
specific for the current application of screening for SARS-CoV-2 via
RT-PCR. For example, cross-reactivity with other coronaviruses would
have violated this assumption. However, cross-reactivity was ruled out
in \cite{KitComparison}. These assumptions lead to our model, which is
illustrated in Figure~\ref{fig:likelihood} and summarized in equation
\eqref{eq:likelihood}.
\begin{align}\label{eq:likelihood}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N (1-\sens)^{\theta_i} =
    \spec(1-\Se)^{\sum_i\theta_i}.
\end{align}

@@

\subsection*{Application: scheme false-negative rate}
We calculate the false-negative rate for a single \emph{infected}
individual, henceforth referred to as "Donald", under our model and
under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Equivalently, we ask: what is the probability of
not identifying Donald as infected under Dorfman pooling?

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a specificity of
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, the increase in the scheme false-negative rate
(compared to $1-\Se$), expressed by $\err{exact}$, is at leaset
$60\%$ (Figure \ref{fig1}).

@@

\section*{Discussion}
In this study, we developed a more realistic and novel probabilistic
model for the outcomes of pooled RT-PCR tests, parameterized for
SARS-CoV-2. Contrary to the common assumption, we assume, based on
data (Table~\ref{table}), that multiple infected individuals increase
the likelihood of a positive pooled test result. A direct consequence
of our model is that low values of infection prevalence increase the
false-negative rates of Dorfman pooling. Importantly, our results
remain qualitatively similar under varying parameter values, in the
observed ranges \cite{KitComparison,EstimatingRatesKucrika,
  EstimatingRatesLourenco, InterpretingCOVID19Test} (Figure
\ref{fig1}). Therefore, our results give rise to a conflict: low
infection prevalence leads to high efficiency of Dorfman pooling
\cite{DorfmanYuvalDor},but also increases false-negative rates.

As the COVID-19 pandemic progresses, the infection prevalence in
various tested populations undergoes frequent changes. Hence, as our
results suggest, pooling schemes employed for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers \cite{JTM}
or presymptomatic and asymptomatic individuals \cite{RobinHood}, and
can be crucial for controlling outbreaks \cite{MinaScience}.

Furthermore, an especially important consideration in designing
pooling schemes is explicitly taking into account the intrinsic RT-PCR
error rates . As we have demonstrated above, analyzing the expected
error rates of pooling schemes under a realistic error model is
readily achievable. Designing pooling schemes that optimize testing
while accounting for error rates is, however, considerably more
involved.  Nonetheless, various algorithms explicitly incorporate
error rates into their pooling optimization processes \cite{Kim,
  OptimalDorfmanPool}, including a recent study by our group which
employed the herein described error model \cite{DOPE}.  In our
proposed algorithm, we have used an experimental design criterion to
create an algorithm which iteratively performs pooling, updating its
information based on previous results. In doing so, our pooling scheme
shows potential in terms of both reducing false-negative and
false-positive rates, as well as increasing efficiency, although it
has not yet been implemented in real-world settings.

>>>>

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is the backbone and gold standard of infection surveillance across the globe, and a key component in the fight against the COVID-19 pandemic. One major application of this technology is the rapid testing of contacts of positive individuals in order to cut transmission chains. Moreover, in some
countries, RT-PCR testing has become a routine procedure of
surveillance, testing incoming air passengers in airports
\cite{PoolingAirports, TestingAirportsJapan}, schools students
\cite{PoolSchool}, or even the entire adult population
\cite{MassPooling}.

As such, the need for large-scale testing has resulted in the
application of techniques to increase the throughput of RT-PCR tests
\cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman, MatrixPooling,
  LionDorfman, CherifReview}. These techniques, often termed
\emph{pooling}, or group testing, have originated in the 1940's in the
seminal work of Dorfman 
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroughly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

@@

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID, CherifReview}. Consequently,
studies assumed that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been assumed identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool,
  CherifReview}, as well as in more specific, SARS-CoV-2 focused
studies \cite{Simplistic1, Simplistic2}. In this study we challenge
this commonly made assumption and demonstrate how using a more accurate
probabilistic model affects the estimation of false-negative rates for
Dorfman pooling.

Our analysis has several implications: Previously estimated
optimal Dorfman pool size \cite{OptimalDorfmanPool} should be
revisited in general and specifically for the case of SARS-CoV-2, as it is influenced by the
prevalence of infection. The error rates of other pooling schemes in
use should also be revisited \cite{BayesianDorfman, Kim}. Moreover, studies that seek precise error
models should be conducted for other pathogens as well, since the
standard error model could prove flawed. Specifically, all assumptions should be scrutinized,
including (e.g.) the independence of tests, an extremely widespread
assumption in the group testing literature \cite{Kim,
  OptimalDorfmanPool} which dates back to Dorfman's original paper
\cite{DorfmanOriginal}. Finally, new methods should be developed in
order to take advantage of more realistic models.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume a more
realistic model: amplification of viral RNA succeeds or fails for each
sample independently. Furthermore, according to \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool either. For lack of
data pointing otherwise, we incorporate this assumption into our model
with a small modification. We do assume that a false amplification can
occur only once per pool. However, we also assume false amplification
is independent of any other correct amplification. Specifically, it is
possible that every correct amplification fails \emph{and} an
erroneous one occurs simultaneously. This assumption is somewhat
specific for the current application of screening for SARS-CoV-2 via
RT-PCR. For example, cross-reactivity with other coronaviruses would
have violated this assumption. However, cross-reactivity was ruled out
in \cite{KitComparison}. These assumptions lead to the following model,
illustrated in Figure~\ref{fig:likelihood} and summarized in equation
\eqref{eq:likelihood}.
\begin{align}\label{eq:likelihood}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N (1-\sens)^{\theta_i} =
    \spec(1-\Se)^{\sum_i\theta_i}.
\end{align}

@@

\subsection*{Application: scheme false-negative rate}
We calculate the false-negative rate for a single \emph{infected}
individual, henceforth referred to as ''Donald'', under our model and
under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Equivalently, we ask: what is the probability of
not identifying Donald as infected under Dorfman pooling?

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We consider a specificity of
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}. We observe that the increase in the scheme false-negative rate
(compared to $1-\Se$), expressed by $\err{exact}$, is at least
$60\%$ (Figure \ref{fig1}).

@@

\section*{Discussion}
In this study a novel probabilistic model is developed to capture realistic features of the outcomes of pooled RT-PCR tests, parameterized for
SARS-CoV-2. Contrary to the common assumption, we assume, based on
data (Table~\ref{table}), that multiple infected individuals increase
the likelihood of a positive pooled test result. A direct consequence
of our model is that low values of infection prevalence increase the
false-negative rates of Dorfman pooling. Importantly, our results
remain qualitatively similar under varying parameter values, in the
observed ranges \cite{KitComparison,EstimatingRatesKucrika,
  EstimatingRatesLourenco, InterpretingCOVID19Test} (Figure
\ref{fig1}). Therefore, our results give rise to a conflict: low
infection prevalence leads to high efficiency of Dorfman pooling
\cite{DorfmanYuvalDor}, but also increases false-negative rates.

As the COVID-19 pandemic progresses, the infection prevalence in
various tested populations undergoes frequent changes. Hence, as our
results suggest, pooling schemes employed for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers \cite{JTM},
nursing homes \cite{DorfmanYuvalDor},
or presymptomatic and asymptomatic individuals \cite{RobinHood}, and
can be crucial for controlling outbreaks \cite{MinaScience}.

Furthermore, an especially important consideration in designing
pooling schemes is explicitly taking into account the intrinsic RT-PCR
error rates. As illustrated above, analyzing the expected
error rates of pooling schemes under a realistic error model is
readily achievable. Designing pooling schemes that optimize testing
while accounting for error rates is, however, considerably more
involved.  Nonetheless, various algorithms explicitly incorporate
error rates into their pooling optimization processes \cite{Kim,
  OptimalDorfmanPool}, including a recent study by our group which
employed the herein described error model \cite{DOPE}.  In our
proposed algorithm, we have used an experimental design criterion to
create an algorithm which iteratively performs pooling, updating its
information based on previous results. In doing so, our pooling scheme
shows potential in terms of both reducing false-negative and
false-positive rates, as well as increasing efficiency, although it
has not yet been implemented in real-world settings.
